---
title: "SDG 2.2 and the 2023 Nutrition Month Data"
format: docx
---

## SDG 2.2

### Target 2.2
By 2030, end all forms of malnutrition, including achieving, by 2025, the internationally agreed targets on stunting and wasting in children under 5 years of age, and address the nutritional needs of adolescent girls, pregnant and lactating women and older persons.  


### Indicators 
**2.2.1**
Prevalence of stunting (height for age <-2 standard deviation from the median of the World Health Organization (WHO) Child Growth Standards) among children under 5 years of age.  
The targets set for 2025 and 2030 for stunting among children under five are to achive a reduction of 40% and 50% reapectively from the levels of 2012.

**2.2.2**
Prevalence of malnutrition (weight for height >+2 or <-2 standard deviation from the median of the WHO Child Growth Standards) among children under 5 years of age, by type (wasting and overweight).  
The targets set for 2025 and 2030 for wasting among children under five are to reduce and maintain the prevalence of wasting to 5% and 3% respectively.  

The last three DHSs conducted in 2000, 2006/7 and 2016 have shown minimal reduction in the prevalence of stunting and wasting among children under 5 (figure 1). In fact, the prevalence of stunting was found to be 17.5% in both 2006/7 and 2016. A 40% reduction (the target for 2025) in stunting from the levels of 2012 translates to 10.5% and a 50% reduction (the target for 2030) translates to 8.8%. The nutrition month data for the last three years - 2021, 2022, and 2023 - reveal a worrying trend. The prevalence of stunting had gone up from 7.4% in 2021 to 10.3% in 2023, still lower than 10.5% the target for 2025, but if the recent upward trend continues it is likely that we will end up with a prevalence of stunting well above the target.  
The prevalence of wasting among children under five is 10% in 2023 which is well above the 2025 target of 5%.  
The DHS data is based on a nationally representative sample and the nutrition month data is based on an attempt to measure all eligible children with an estimated participation rate of 96.5%. Therefore, the differences in the rates are unlikely to be due to the differences in the methodology.  
The nutrition month data includes prevalence by age groups - under 1 year, 1 to 2 years and 3 to 5 years. The prevalence of stunting has increased in all three age groups (@fig-stunt)
```{r}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
library(gridExtra)
# library(cowplot)

# Enter data
ntmnth_age <- tribble(
  ~agegp, ~year, ~stunt, ~waste, ~underwt,
  "Infants", 2023, 6.9, 5.0, 10.5,
  "1-2 years", 2023, 12.0, 8.4, 15.4,
  "3-5 years", 2023, 10.8, 12.0, 19.5,
  "Infants", 2022, 6.2, 5.8, 9.7,
  "1-2 years", 2022, 10.8, 8.5, 13.8,
  "3-5 years", 2022, 9.5, 12.0, 17.5,
  "Infants", 2021, 4.7, 4.7, 7.2,
  "1-2 years", 2021, 8.2, 6.9, 10.4,
  "3-5 years", 2021, 8.0, 9.7, 14.4
)
# Order age groups - infant > 1-2 > 3-5 

ntmnth_age <- ntmnth_age %>% 
  mutate(agegp = factor(agegp)) %>% 
  mutate(agegp=fct_relevel(agegp,c("Infants", "1-2 years", "3-5 years"))) 

my_bar <- function(x){ggplot(data = ntmnth_age, 
                             mapping = aes(x = year, y = {{x}}, fill = agegp)) + 
    geom_bar(stat = "identity", position=position_dodge())+
    geom_text(aes(label= {{x}}), vjust= -0.2, color = "black",
              position = position_dodge(0.9), size=5)}

dhs_o <- tribble(
  ~year,  ~underwt, ~stunt, ~waste,
  1993, 33.8, 29.7, 17.5,
  1996, 29.3, 26.1, 15.3,
  2000, 22.8, 18.4, 15.5,
  2007, 21.1, 17.3, 14.7,
  2016, 20.5, 17.3, 15.1
)

dhs_l <- dhs_o %>% 
  pivot_longer(
    cols = 2:4, 
    names_to = "ind", 
    values_to = "Percent")

dhs_sdg <- dhs_o %>% filter(year>1999) %>% 
  pivot_longer(
    cols = 3:4, 
    names_to = "ind", 
    values_to = "Percent")

```

```{r}
#| echo: false
#| message: false
#| label: fig-stunt
#| fig-cap: "Stunting among under 5"
my_bar(stunt) + 
  scale_fill_grey() + 
  theme_light(base_size = 16) + 
  labs(x = "", y = "Percent", fill = "", title = "") +
  scale_y_continuous(limits = c(0, 14)) +
  scale_color_discrete(name = "") + 
  theme(legend.position = "bottom") + 
  scale_fill_manual("", values = c("Infants" = "#9ECAE1", "1-2 years" = "#2171B5", "3-5 years" = "#08306B"))

```

The prevalence of wasting has increased in all three age groups and there is a clear increase in wasting with increasing age (@fig-waste)


```{r}
#| echo: false
#| message: false
#| label: fig-waste
#| fig-cap: "Wasting among under 5"
my_bar(waste) + 
  scale_fill_grey() + 
  theme_light(base_size = 16) + 
  labs(x = "", y = "Percent", fill = "", title = "") +
  scale_y_continuous(limits = c(0, 14)) +
  scale_color_discrete(name = "") + 
  theme(legend.position = "bottom") +
  scale_fill_manual("", values = c("Infants" = "#9ECAE1", "1-2 years" = "#2171B5", "3-5 years" = "#08306B"))

```
It is a good idea to look at the overall trends over the past couple of decades but we don't have nutrition month data going back so many years. We do have DHS data for 2000, 2006/7 and 2016 on the prevalence of stunting and wasting among children under 5 years. The DHS data and the nutrition month data are strictly not comparable. The DHS data is based on a nationally representative sample and the nutrition month data is from measurements made on all eligible children.  
There had been a slight drop in stunting and wasting between 2000 and 2006/7 but the rates remained static between 2006/7 and 2016.

```{r}
#| echo: false
#| message: false
#| label: fig-dhs
#| fig-cap: "Stunting and Wasting among Cildren Under 5 (DHS 2000, 2006/7 & 2016)"
ggplot(data = dhs_sdg, mapping = aes(x = year, y = Percent, group = ind)) +
theme_light(base_size = 16) +   
geom_line(aes(colour = ind)) +
  labs(x = "", y = "Percent", fill = "", title = "") +
geom_point(aes(colour = ind)) +
scale_x_continuous(breaks= c(2000, 2007, 2016)) +
scale_y_continuous(limits = c(0, 20)) +
scale_color_manual("", values=c("#CC6666", "#9999CC"),
breaks = c("stunt", "waste"),
labels = c("Stunted", "Wasted")) +
theme(legend.position = "bottom")
```

There is a big difference in the prevalence of stunting and wasting between 2016 DHS and 2021 Nutrition month.

```{r}
#| echo: false
#| message: false
#| label: fig-ntmth
#| fig-cap: "Stunting and Wasting among Cildren Under 5 (Nutrition months 2021, 2022 & 2023)"

ntmnth_o <- tribble(
  ~year,  ~underwt, ~stunt, ~waste,
  2021, 12.2, 7.4, 8.2,
  2022, 15.3, 9.2, 10.1,
  2023, 17.1, 10.3, 10.0
)

ntmnth_sdg <- ntmnth_o %>% 
  pivot_longer(
    cols = 3:4, 
    names_to = "ind", 
    values_to = "Percent")
ntmnth_sdg <- ntmnth_sdg %>%
mutate(ind = factor(ind)) %>%
mutate(ind=fct_relevel(ind,c("stunt", "waste")))
ggplot(data = ntmnth_sdg, mapping = aes(x = year, y = Percent, group = ind)) +
geom_line(aes(colour = ind)) +
geom_point(aes(colour = ind)) +
    theme_light(base_size = 16) + 
  labs(x = "", y = "Percent", fill = "", title = "") +
scale_x_continuous(breaks= c(2021, 2022, 2023))+
scale_y_continuous(limits = c(0, 20)) +
  scale_color_manual(values=c("#CC6666", "#9999CC"),
name = "",
breaks = c("stunt", "waste"),
labels = c("Stunted", "Wasted")) +
theme(legend.position = "bottom")
```


